\large{Vipassana for Hackers} \\ \Huge{The Proposal} \\ \large\textit{Version 0.1}

### 2019-12-04 the-proposal/the-proposal.tex
proposal: study design

``I can't think of a single discovery of disease which had more than one initial
sample. .... You can't do statistical analysis of an initial discovery.''
\cite{ramachandran2019relevance}

@@

A lifelong study of a single, increasingly-experienced Vipassana meditator
is, in essence, an exploratory process of discovering what researchers might be
bothered to study in a (single) blind trial with active and inactive controls, over a
much larger sample. This process could be described as xxx

\todo{(Harari) / (Altered Traits) quotes}

Research tool initially, and research tool + research subject later on.

@@

\subsection{Study Design / Research Method}
- case study type?

>>>>

\begin{quotation}
  ``I can't think of a single discovery of disease which had more than one initial
  sample. ... You can't do statistical analysis of an initial discovery.''
  \cite{ramachandran2019relevance}
\end{quotation}

@@

A lifelong study of a single, increasingly-experienced Vipassana meditator
is, in essence, an exploratory process of discovering what researchers might be
bothered to study in a controlled trial with both active and inactive controls and
(single) blinding, over a much larger sample. This exploratory process has been
described by Goleman and Davidson in \textit{Altered Traits}:

\begin{quotation}
  ``Perhaps one day an ultralong study will give us the equivalent of video on how
  altered traits emerge. For now, as the Brewer group conjectured, meditation seems
  to transform the resting state---the brain's default mode---to resemble the
  meditative state.

  Or, as we put it long ago, the after is the before for the next
  during.'' \cite{goleman2017altered}
\end{quotation}

The demands of such an ``ultralong study'' were echoed by Harari in \textit{21
  Lessons for the 21st Century}:

\begin{quotation}
  ``Some universities and laboratories have indeed begun using meditation as a
  research tool rather than as a mere object for brain studies. Yet this process is
  still in its infancy, partly because it requires an extraordinary investment on the
  part of the researchers.'' \cite{harari201821}
\end{quotation}

The line between ``mere object for brain studies'' and ``research tool'' begins to
blur when we consider using meditation for both simultaneously. However, it seems remiss to neglect
gathering whatever objecting EEG/fMRI/MEG data from the long-term subject-researcher
when that data is so readily available.

@@

The study proposed is an ultralong (lifelong) study of an individual --- myself. As
far as quantitative data gathered from brain scans is concerned, I am effectively
making myself available as a human guinea pig. As far as qualitative data is
concerned, research will take the form of a deep case study. As I will represent
Ramachandran's ``initial sample'', statistical analysis will not be relevant. If the
study proves productive, future studies targeting a specific hypothesis about
Vipassana in a large sample can use learnings from this study in their statistical
structure.

\subsection{Study Design}

\subsubsection{Qualitative}

Qualitative research into the specific nature of human consciousness or the broader
nature of consciousness as it applies to any organism or artificial intelligence can
only be performed directly. In this case, a case study performed by an individual
(myself) will explore the question of consciousness and the mind-body problem
directly, through Vipassana meditation, for a prolonged period of time (the remainder
of my life). Many theories and questions currently exist regarding
consciousness, such as \textit{Theory of Mind}, \textit{Theory of Panpsychism}, and
the \textit{Simulation Hypothesis}. This study will not address any one of them
directly, nor will it absorb any specific hypothesis. The data recovered from
practicing Vipassana meditation directly for prolonged periods of time will almost
certainly overlap with many such ideas but themes and narratives are much more likely
to be emergent, rather than conforming to an existing hypothesis of consciousness.

Phenomenology, in the broadest sense, may be used to describe experiences as they
pertain to a lifelong trajectory of altered traits. Phenomenological descriptions of
deep meditative states, or even of altered traits, is not the goal,
however. If other narrative tools emerge over the course of the study, I will augment
phenomenological descriptions and imagery with those tools.

Vipassana's instruction provides us with claims we can evaluate: ``[Vipassana] explores the entire field of
mind and matter'' \cite{goenka1999discourses}, ``[Vipassana] is a technique that will
eradicate suffering'' \cite{dhammaorg2014code}, and ``Vipassana aims at ... total
liberation and full enlightenment'' \cite{dhammaorg2014code}.

\subsubsection{Quantitative}

Measurements performed during the lifelong qualitative study
